---
title: "Flies Survival Analysis"
author: "Luis Gracia <lgraval@gmail.com>"
date: "May 6, 2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

# Setup

## Install packages

```{r eval=FALSE, include=FALSE}
install.packages(c("knitr", "tidyverse", "tidyr", "dplyr", "readxl", "janitor", "survival", "survminer", "data.table"))
```

## Notebook general settings

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Needed libraries

```{r include=FALSE}
library(tidyr)
library(dplyr)
library(readxl)
library(janitor)
library(survival)
library(survminer)
library(data.table)
library(tools)
```

# Prepare data

## Set input and output files

```{r}
inputfile = "tests.xlsx"
worksheet = "dead+escaped+unaccounted+final"
skip_unaccounted = FALSE        # TRUE/FALSE
```

## Import data

Load data from Excel and report a sample:

```{r}
input = read_excel(inputfile, sheet=worksheet)
sample_n(input, 5)
```

# Transform data

First reshape and clean data:
1. Reshape the date column into rows (removing blanks).
2. Identify values marked with ":".
2. Split dead and escaped events marked with "+".
3. Set correct formats for group, tube, dead, escaped and date columns,
4. Calculating day passed per group/tube.

TODO:
- Remove flies0
- Manage initial dates marked with ":"

```{r}
data0 = input %>%
  gather(date, flies0, -tube, -group, na.rm=T) %>%
  separate(flies0, c("marked", "flies"), ":", remove=T, fill="left") %>%
  separate(flies, c("dead", "escaped"), "[+]", fill="right") %>%
  mutate(group = as.factor(group),
         tube = as.factor(tube),
         dead = as.numeric(dead),
         escaped = as.numeric(escaped),
         date = excel_numeric_to_date(as.numeric(date))) %>%
  group_by(group, tube) %>% mutate(days = as.integer(date - min(date, na.rm=T)))
```

Let's now split the data in three datasets:
- Initial flies per group/tube.
- Final flies per group/tube.
- Journal of count of events per day.

```{r}
initial = data0 %>% filter(days == 0) %>% select(-marked, -escaped) %>% rename(initial = dead)
final = data0 %>% filter(!is.na(marked)) %>% select(-marked, -escaped) %>% rename(final = dead)
journal = data0 %>% filter(days != 0) %>% filter(is.na(marked)) %>% select(-marked)

summary(journal)
```

Create summary tally:
- By group/tube for calculating unaccounted flies per tube
- By group for inspection

```{r}
tally = initial %>% select(-date, -days) %>%
  full_join(
    journal %>% summarise(dead = sum(dead, na.rm=T), escaped = sum(escaped, na.rm=T)),
    by = c("group", "tube")
  ) %>%
  full_join(
    final %>% select(-date, -days),
    by = c("group", "tube")
  ) %>%
  mutate_all(~ if_else(is.na(as.numeric(.x)), 0, as.numeric(.x))) %>%
  mutate(unaccounted = abs((final + dead + escaped) - initial))

tally

bind_rows(
  tally %>% group_by(group) %>% summarise_if(is.numeric, sum),
  tally %>% group_by() %>% summarise_if(is.numeric, sum)
)
```

The total number of events should be:

```{r}
expected_events = tally %>% mutate(events = (dead + escaped + unaccounted)) %>% group_by() %>% summarise_if(is.numeric, sum) %>% select(events)
```

Construct events dataset for use in survival analysis packages by expanding rows with dead and escaped (censored) events. Additionally add censored events for flies unaccounted for.

```{r}
events = bind_rows(
  journal %>%                                       # dead flies: 1
    select(-escaped) %>%
    filter(dead > 0) %>%
    uncount(dead, .id="fly") %>%
    mutate(event=as.integer(1)),
  journal %>%                                       # escape flies: 0
    select(-dead) %>%
    filter(!is.na(escaped)) %>%
    uncount(escaped, .id="fly") %>%
    mutate(event=as.integer(0)),
  tally %>%                                         # unaccounted flies: 0  (-1 if skip_unaccounted=TRUE)
    select(group, tube, unaccounted) %>%
    filter(unaccounted > 0) %>%
    uncount(unaccounted, .id="fly") %>%
    mutate(days=1, event=ifelse(skip_unaccounted, -1, 0), date=as.Date(NA))
)

events = events %>% filter(event == 0 | event == 1)

events
```

# Analyze

## Export to Prism

```{r echo=FALSE}
prismfile = paste(paste(file_path_sans_ext(inputfile), worksheet, "prism", sep="_"), ".csv", sep="")
```

Export to Prism in file *`r prismfile`*.

```{r}
prism = events %>%
  group_by(group) %>%
  mutate(id = 1:n()) %>%
  dcast(group+tube+days+fly+event+id ~ group, value.var="event", fill="") %>%
  select(-group, -tube, -fly, -event, -id)

write.csv(prism, prismfile)
```

## Suvival Analysis

First, create a survival object and fit by group and output the median and CI intervals:

```{r}
survival = survfit(Surv(days, event) ~ group, data=events)
surv_median(survival)
```

And create a pretty graph:

```{r}
ggsurvplot(survival, data=events, xlab="Days", palette="jco")
ggsurvplot(survival, data=events, xlab="Days", pval=T, conf.int=T, surv.median.line="hv", palette="jco")
```

We can also test for differences:
```{r}
survdiff(Surv(days, event) ~ group, data=events)
```